---
title: "Lab 10: Distances and the Border Zone"
subtitle: 'Ecosystem Science and Sustainability 330'
author:
  - name: Chippy Marx
    email: mchippy@colostate.edu
format: html
execute:
  echo: true
---

Load libraries

```{r}
# spatial data science

library(tidyverse)
library(sf)
library(units)
library(dplyr)
library(flextable)
library(tigris)
library(maps)
library(flextable)

# Data
library(AOI)
library(rnaturalearth)
library(rnaturalearthhires)

# Visualization
library(gghighlight)
library(ggrepel)
library(knitr)
```

# Question 1

```{r}
#equal area projection 
eqdc <- 5070

aoi_get(country = c("MX", "CA", "USA"))

#load in city data 
uscities <- read_csv("data/uscities.csv")

#explore city data 

head(uscities)

# Explore US cities data 
str(uscities)
summary(uscities)
names(uscities)

# Convert to spatial object 
uscities_sf <- uscities %>%
  st_as_sf(coords = c("lng", "lat"), crs = 4326, remove = FALSE)

#filter out the states we don't need 
uscities_border <- uscities_sf %>%
  filter(!state_name %in% c("Alaska", "Hawaii", "Puerto Rico"))

#reproject for distance measurements 
uscities_border_proj <- st_transform(uscities_border, crs = 5070) 


```

# Question 2

**Create Resolved State Boundaries**

```{r}
#make data frame with us state names and their centers 
df <- data.frame(name = state.name, 
                 X = state.center$x,
                 Y = state.center$y)

#turn data frame into sparial object usinf the x/y coordinates 
cities <- st_as_sf(df, coords = c("X", "Y"), crs = 4326)


states <- USAboundaries::us_states(resolution = "low") |>
  filter(!state_abbr %in% c("HI", "AK", "PR")) |>
  st_transform(eqdc)

#merge all states into one polygon
us_union <- st_union(states)

#get just the boundary of the states 
us_boundaries <- st_boundary(us_union)

#project city points into right coordinate system 
cities_proj <- st_transform(cities, crs = eqdc)

#make map of boundaries 
cities_proj <- uscities_border_proj  
us_boundaries <- st_cast(us_boundaries, "MULTILINESTRING")
plot(us_boundaries, col = "black", lwd = 0.5, main = "Resolved State Boundaries")


```

**Create boundaries and Calculate City Distances to them.**

*I calculated and added all the distances needed at once and then will create the tables below*

```{r}

#create state borders 
state_boundaries <- st_combine(states) |>
  st_boundary() |>
  st_cast("MULTILINESTRING")

#Downladed the North America data, then made it spatial 
north_america <- ne_countries(continent = "North America", returnclass = "sf") |>
  st_transform(crs = eqdc)

#define borders 

#mexico
mexico <- north_america |> filter(admin == "Mexico")

#canada 
canada <- north_america |> filter(admin == "Canada")

#find boundaries of the countries , make a multistring 

mex_border <- st_boundary(mexico) |> st_cast("MULTILINESTRING")

can_border <- st_boundary(canada) |> st_cast("MULTILINESTRING")


#add the distances to the borders we are interested to the city data frame
cities_with_distances <- uscities_border_proj |>
  mutate(
    dist_to_national_km = st_distance(geometry, us_boundaries) |> set_units("km") |> drop_units(),
    dist_to_state_km    = st_distance(geometry, state_boundaries) |> set_units("km") |> drop_units(),
    dist_to_mexico_km   = st_distance(geometry, mex_border) |> set_units("km") |> drop_units(),
    dist_to_canada_km   = st_distance(geometry, can_border) |> set_units("km") |> drop_units()
  )


```

**2.1: Distance to US border/coastline**

```{r}
# Top 5 cities farthest from the US  border
top5_farthest_us <- cities_with_distances %>%
  st_drop_geometry() %>%
  select(city, state_name, dist_to_national_km) %>%
  rename(
    State = state_name,
    `distance to US border (km)` = dist_to_national_km
  ) %>%
  arrange(desc(`distance to US border (km)`)) %>%
  slice(1:5)

flextable(top5_farthest_us)


flextable(top5_farthest_us)
```

**2.2: Distances to state border**

```{r}
top5_farthest_state <- cities_with_distances %>%
  st_drop_geometry() %>%
  select(city, state_name, dist_to_state_km) %>%
  rename(
    state = state_name,
    `distance to state border (km)` = dist_to_state_km
  ) %>%
  arrange(desc(`distance to state border (km)`)) %>%
  slice(1:5)

# Show as flextable
flextable(top5_farthest_state)


# 4. Show as flextable
flextable(top5_farthest_state)
```

**2.3: Distance to Mexico**

```{r}
# Top 5 cities farthest from the **Mexican border**
top5_farthest_mexico <- cities_with_distances %>%
  st_drop_geometry() %>%
  select(city, state_name, dist_to_mexico_km) %>%
  rename(
    state = state_name,
    'distance to Mexico border (km)'= dist_to_mexico_km
  ) %>% 
  arrange(desc(`distance to Mexico border (km)`)) %>%
  slice(1:5)

flextable(top5_farthest_mexico)
```

**2.4: Distance to Canada**

```{r}
# Top 5 cities farthest from the Canadian border

top5_farthest_canada <- cities_with_distances %>%
  st_drop_geometry() %>%
  select(city, state_name, dist_to_canada_km) %>%
  rename(
    state = state_name, 
    'distance to Canadian border (km)'= dist_to_canada_km
  ) %>% 
  arrange(desc(`distance to Canadian border (km)`)) %>%
  slice(1:5)

flextable(top5_farthest_canada)
```

# Question 3

**3.1: Data**

```{r}
# The 10 biggest cities by population
top10_cities <- uscities_sf %>%
  arrange(desc(population)) %>%
  slice(1:10)

#Make the figure 
ggplot() +
  geom_sf(data = north_america, fill = "lightgray", color = "black", lty = 1, size = 0.5) +
  geom_sf(data = us_union, fill = NA, color = "darkblue", lty = 2, size = 1) +
  geom_sf(data = states, fill = NA, color = "black", lty = 3, size = 0.3) +
  geom_sf(data = top10_cities, color = "red", size = 2) +
  ggrepel::geom_label_repel(
    data = top10_cities,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates",
    size = 3,
    min.segment.length = 0
  ) +
  theme_minimal() +
   ggtitle("Top 10 Largest US Cities with North America and State Borders") +
  theme(plot.title = element_text(hjust = 0.5))

```

**3.2: City Distance from border**

```{r}
highlight_cities <- cities_with_distances %>%
  arrange(desc(dist_to_national_km)) %>%
  slice(1:5)

ggplot() +
  geom_sf(data = cities_with_distances, aes(color = dist_to_national_km), size = 1) +
  geom_sf(data = highlight_cities, color = "red", size = 2) + 
  geom_label_repel(
    data = highlight_cities,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates",
    size = 3
  ) +
  scale_color_viridis_c(name = "Distance from National Border (km)") +
  theme_minimal() +
  ggtitle("Cities Colored by Distance from U.S. National Border") +
  theme(plot.title = element_text(hjust = 0.5))


```

**3.3: City Distance to nearest state**

```{r}
# Match by city + state to avoid duplicates, I found that many of the cities had the same names
# Create a city_state_id column for matching
cities_with_distances <- cities_with_distances %>%
  mutate(city_state_id = paste0(city, ", ", state_name))

top5_farthest_state <- top5_farthest_state %>%
  mutate(city_state_id = paste0(city, ", ", state))

# Match 
highlight_state_cities <- cities_with_distances %>%
  filter(city_state_id %in% top5_farthest_state$city_state_id)


# Make the map
ggplot() +
  geom_sf(data = cities_with_distances, aes(color = dist_to_state_km), size = 1) +
  geom_sf(data = highlight_state_cities, color = "red", size = 2) +
  geom_label_repel(
    data = highlight_state_cities,
    aes(label = paste0(city, ", ", state_name), geometry = geometry),
    stat = "sf_coordinates",
    size = 3
  ) +
  scale_color_viridis_c(name = "Distance from State Border (km)") +
  theme_minimal() +
  ggtitle("Cities Colored by Distance from Nearest State Border") +
  theme(plot.title = element_text(hjust = 0.5))

```

**3.4: Equidistance boundary from Mexico and Canada**

```{r}
# Create absolute difference column
cities_with_distances <- cities_with_distances %>%
  mutate(diff_mex_can = abs(dist_to_mexico_km - dist_to_canada_km))

#  Filter cities within 100 km difference
equidistant_cities <- cities_with_distances %>%
  filter(diff_mex_can <= 100)

#Get 5 most populous cities in that zone
top5_equidistant <- equidistant_cities %>%
  arrange(desc(population)) %>%
  slice(1:5)

#Plot
ggplot(data = cities_with_distances) +
  geom_sf(aes(color = diff_mex_can), size = 1) +
  gghighlight(diff_mex_can <= 100, use_direct_label = FALSE) +
  geom_sf(data = top5_equidistant, color = "red", size = 2) +
  geom_label_repel(
    data = top5_equidistant,
    aes(label = paste0(city, ", ", state_name), geometry = geometry),
    stat = "sf_coordinates",
    size = 3
  ) +
  scale_color_viridis_c(name = "Abs(Mex - Canada Distance) km") +
  theme_minimal() +
  ggtitle("Cities Approximately Equidistant from Mexico and Canada (±100 km)") +
  theme(plot.title = element_text(hjust = 0.5))

```

# Question 4

**4.1: Quantifying border zone**

```{r}
# add column with distance to border
cities_with_distances <- cities_with_distances %>%
  mutate(in_border_zone = dist_to_national_km <= 160)

# how many cities are in the border zone
n_cities_border <- sum(cities_with_distances$in_border_zone)

# Total people in that zone
pop_border <- cities_with_distances %>%
  filter(in_border_zone) %>%
  summarise(total_population = sum(population, na.rm = TRUE)) %>%
  pull(total_population)

# sum total US population
pop_total <- sum(cities_with_distances$population, na.rm = TRUE)

#Percent of population in border zone
pct_border <- round((pop_border / pop_total) * 100, 1)

#Create table
border_stats <- data.frame(
  Metric = c(
    "Cities within 100 miles of U.S. border",
    "People living in those cities",
    "Percent of total U.S. city population"
  ),
  Value = c(
    n_cities_border,
    format(pop_border, big.mark = ","),
    paste0(pct_border, "%")
  )
)

#make table 
library(flextable)
flextable(border_stats)

```

*Its even more then the article said! The article said that about 200 million people live in this zone*

**4.2: Mapping border zone**

```{r}
#Filter 10 most populous cities in border zone
top10_border <- cities_with_distances %>%
  filter(in_border_zone) %>%
  arrange(desc(population)) %>%
  slice(1:10)

# Map it 
ggplot(data = cities_with_distances) +
  geom_sf(aes(color = population), size = 1) +
  gghighlight(in_border_zone, use_direct_label = FALSE) +
  scale_color_gradient(low = "orange", high = "darkred", name = "Population") +
  geom_label_repel(
    data = top10_border,
    aes(label = paste0(city, ", ", state_name), geometry = geometry),
    stat = "sf_coordinates",
    size = 3
  ) +
  theme_minimal() +
  ggtitle("Cities Within 100 Miles of U.S. National Border") +
  theme(plot.title = element_text(hjust = 0.5))

```

**4.3: label by the most populous city in each state in the danger zone**

```{r}
# Most populous city per state within 100-mile border zone
top_city_per_state_border <- cities_with_distances %>%
  filter(in_border_zone) %>%
  group_by(state_name) %>%
  slice_max(order_by = population, n = 1, with_ties = FALSE) %>%
  ungroup()

# 2. Map
ggplot(data = cities_with_distances) +
  geom_sf(aes(color = population), size = 1) +
  gghighlight(in_border_zone, use_direct_label = FALSE) +
  scale_color_gradient(low = "orange", high = "darkred", name = "Population") +
  geom_label_repel(
    data = top_city_per_state_border,
    aes(label = paste0(city, ", ", state_name), geometry = geometry),
    stat = "sf_coordinates",
    size = 3
  ) +
  theme_minimal() +
  ggtitle("Most Populous City per State Within 100-Mile Border Zone") +
  theme(plot.title = element_text(hjust = 0.5))

```

You have to click on figure to see all the states labeled.
